---
title: "Quarto Assessment MT24"
format: 
  html: 
    embed-resources: true
editor: source
---

# Introduction

The aim of this analysis is to assess whether drought has an impact on the evolution of beak morphology in the medium ground finch, *Geospiza fortis*.

Beak morphology is believed to evolve in response to availability of food and preferred food of a species, in part defined by the niche a species occupies and in part by environmental effects altering what foods are available and in what quantity.
Freak events like droughts can create large selection pressures which act on beak morphology, causing evolution in response to changing food sources.
In 1977, Daphne Major (an island in the Galapagos archipelago, off the coast of Ecuador) experienced a major drought.
This caused changes to and severely limited the food supply on the island, making it such that only large, tough seeds were available for the birds to eat (Boag and Grant, 1981).

Data were collected on beak depth on *G. fortis* before and after the drought; this analysis will examine these data to assess whether there is evidence of the evolution of beak morphology (using beak depth) in response to drought.

```{r Loading packages}
#Loading packages needed for this analysis.
library(Sleuth3) 
library(ggplot2) 
library(tinytex) 
library(arm) 
library(rmarkdown)
library(here)
library(renv)
library(dplyr)
library(multcompView)
library(ragg)
library(patchwork)
```

```{r Investigating data}
#Investigating data.
head(case0201)
str(case0201)
summary(case0201)
colnames(case0201)

#Preserving raw data as .csv file to keep it safe in event of mutation, and making new data frame which can be modified as necessary.
write.csv(case0201, here("data/finches_raw.csv"))
finches_raw <- read.csv(here("data/finches_raw.csv"))

#Factorising "Year" variable.
finches_raw$Year = as.factor(finches_raw$Year)
str(finches_raw)

#Histograms to check spread of data- transformations unsuccessful so will simply use untransformed data.
ggplot(data = finches_raw, aes(x = Depth)) + geom_histogram()
log_depth <- log10(finches_raw$Depth)
ggplot(data = finches_raw, aes(x = log_depth)) + geom_histogram()
sqrt_depth <- sqrt(finches_raw$Depth)
ggplot(data = finches_raw, aes(x = sqrt_depth)) + geom_histogram()

#Setting own colours for each year (ensures colour blind friendly and reproducible).
year_colours <- c("1976" = "magenta2", "1978" = "darkorange")
         
#Plotting exploratory boxplot to see what data look like. Added jitter to show data points, with random seed to ensure jitter is the same every time.
finches_exploratory_figure <- ggplot(data = finches_raw, aes(x=Year, y=Depth, group = Year)) + geom_boxplot(aes(color = Year)) + geom_jitter(aes(color = Year), alpha = 0.3, position = position_jitter(width = 0.2, seed=0)) + labs (x= "Year", y = "Beak depth (mm)") + scale_colour_manual(values = year_colours)

finches_exploratory_figure

#Saving exploratory figure to figures folder.
agg_png("figures/finches_exploratory_figure.png", 
        width = 25,
        height = 15,
        units = "cm",
        res = 300,
        scaling = 1.125)

print(finches_exploratory_figure)
dev.off()
```

# Methods

To analyse the effect of year on beak depth (mm), a one-way ANOVA test will be conducted.
This is because there are two distinct populations, and there is a continuous response variable and a categorical explanatory variable.

The null hypothesis is that the difference in mean beak depth between years is equal to 0.

The alternative hypothesis is that the mean difference in beak depth between years is not equal to 0.

While the assumption of independent samples is violated in this analysis, there is no way to rectify this.
The reason the assumption is violated is that because *G. fortis* could not breed during the 1977 drought, any finches caught in 1978 would also have been alive in 1976 and may have also been included in data from that year; the data from 1978 is a subset of the data from 1976 as only finches alive in 1976 and which survived the 1977 drought were alive in 1978 (Boag and Grant, 1981).

```{r Statistical analysis using lm function}

#Creating linear model to test effect of year on beak depth (mm).
finches_model <- lm(Depth ~ Year, data = finches_raw)

#Creating plots to ensure assumptions of one-way ANOVA test are not violated.
plot1 <- plot(finches_model, which=2)
plot2 <- plot(finches_model, which=1)

combined_assumptions_plot <- (plot1 | plot2)
combined_assumptions_plot

#Running summary of model to assess adjusted R-squared.
summary(finches_model)

#Conducting ANOVA on model to establish if there is a significant effect of year on beak depth. Produces anova table.
finch_anova <- aov(finches_model)
summary(finch_anova)

#Conducting Tukey test to ascertain which groups are different (this is already known but necessary for subsequent code). Produces Tukey post-hoc multiple comparisons table.
finch_tukey <- TukeyHSD(finch_anova)
print(finch_tukey)

#Creates compact letter display, useful for graphically representing significant differences between groups.
finch_cld <- multcompLetters4(finch_anova, finch_tukey)
print(finch_cld)

#Creating new table containing information on mean beak depth in each year, the 3rd quantile of each group and what compact letter each group belongs to.
finch_analysed <- group_by(finches_raw, Year) %>% 
  summarise(mean = mean(Depth), quant = quantile(Depth, probs = 0.75)) %>% 
  arrange(desc(mean))

finch_cld <- as.data.frame.list(finch_cld$Year)
finch_analysed$cld <- finch_cld$Letters

print(finch_analysed)

means <- aggregate(Depth ~ Year, finches_raw, mean)
```

# Results

This one-way ANOVA analysis shows there is a statistically significant difference in mean beak depth between 1976 and 1978.
In 1976, mean beak depth was 9.470mm but in 1978 the mean increased to 10.138mm.

From the summary table for the finch_model, only 10.16% of variation in beak depth is explained by year (adjusted R-squared = 0.1016).
This is surprising as it was expected that the year would play a large role in explaining variation in beak depth, and suggests there may be other factors influencing this.
That said, the F-value for the ANOVA is 21, which is greater than 1 and therefore shows a significant difference between mean beak depth in 1976 and in 1978.
The p-value for this analysis was 8.65\*10\^-6.

The Tukey analysis showed the mean value of beak depth (mm) differed between 1976 and 1978 by 0.669mm.
The p-value is 8.6\*10\^-6, the 95% CI is 0.381 to 0.956mm difference.
As this 95% CI does not include 0, this shows there is a significant difference between groups.

On the boxplot showing the results of this analysis, the median of each group is represented by the bar across the box; the mean of each group is represented by the dark blue dot.
The significant difference between the mean beak depths in each year is shown on the results boxplot, with the compact letter display showing a difference between years.
On the graph, there are two groups, a and b, which correspond to 1976 (b) and 1978 (a).
This means the null hypothesis that there is no difference in mean beak depth between years can be rejected.

```{r Plotting results figure}

#Plotting results figure incorporating compact letter display gained from analysis. Code is similar to that for exploratory figure but have added code to show mean of each group and compact letter display for each group on graph.
finches_results_figure <- ggplot(data = finches_raw, aes(x=Year, y=Depth, group = Year)) + geom_boxplot(aes(color = Year)) + geom_jitter(aes(color = Year), alpha = 0.3, position = position_jitter(width = 0.2, seed=0)) + labs (x= "Year", y = "Beak depth (mm)") + scale_colour_manual(values = year_colours) + stat_summary(fun=mean, colour = "darkblue", geom = "pointrange", shape = 16, size = 0.3, show.legend = FALSE) + geom_text(data = finch_analysed, aes(x=Year, y= quant, label = cld), size = 5, vjust = -1, hjust = -6)

finches_results_figure

#Saving results figure in figures folder.
agg_png("figures/finches_results_figure.png", 
        width = 25,
        height = 15,
        units = "cm",
        res = 300,
        scaling = 1.125)
print(finches_results_figure)
dev.off()

```

# Conclusion

In conclusion, it has been shown that there is a difference in the mean beak depth of the medium ground finch between 1976 and 1978.
In 1976, mean beak depth was 9.470mm but in 1978 the mean increased to 10.138mm.
The results of the one-way ANOVA and Tukey test confirm that there is a statistically significant difference in mean beak depth between 1976 and 1978, coinciding with the 1977 drought.
This therefore has shown that the 1977 drought caused a change in beak morphology in *G. fortis* in response to selection pressures exerted by drought and changing food availability, reflecting the restriction of food supply to large, tough seeds.

## References:

Boag, P. & Grant, P., 1981.
Intense natural selection in a population of Darwin's finches (Geospizinae) in the Galapagos.
*Science,* Volume 214, pp. 82-85.
